---
title: "Basics of R"
output: learnr::tutorial
markdown_extensions:
    - admonition  
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library("lifecontingencies")
tutorial_options(exercise.timelimit = 60)
knitr::opts_knit$set(root.dir = getwd()) 
```

## Introduction

In this tutorial we aim to provide an introduction to the basic capabilities of the `R` software.

`R` is the implementation of the computer language `S`, and was created for statistical analysis. A brief summary may be found by clicking on the following web link to the [Wikipedia page for R](https://en.wikipedia.org/wiki/R_(programming_language)){target="_blank"}. 

If you are already familiar with `R` feel free to skip this tutorial and go directly to the tutorial on Life tables with the package `lifecontingencies`.


## Basic commands
One of the simplest uses of R is as a simple arithmetic calculator as shown here.
 
```{r}
x <- 3
x
sqrt(x)
```

In this example a variable `x` has been declared and given the value $3$. The third input line asks to calculate the square root of `x`; from this you can see that `R` has some built in functions.

We can multiply two variables together, and assign the result to a third:

```{r}
y = 7
z = x * y
z
```

The code above also illustrates that value can also be assigned using the `=` sign. However, from now on and for the sake of consistency, we will always use `<-` for variable assignments.



To see if two variable have the same value, you use the double equals (with
no space between them) `==`. Logical expressions like `x == y` can take `TRUE` or `FALSE` values.

```{r}
x == y
x == y - 2
```

### Exercises

Solve the following exercises using the code chunks below.

1. Create a variable $d=\sqrt[a]{b^c}$ with $a=3$, $b=2$, and $c=9$. 

```{r basic1, exercise=TRUE, exercise.lines = 10, exercise.setup = "prepare-basic"}
a <- 3
b <- 2
c <- 9
# type down your codes below and click run codes to check the answer
# click Hints for additional hints and solutions


```

```{r basic1-hint-1}
# Calculating the cube root is equivalent to raising the variable to the the power of 1/3
## The solutions are included in the next hint page
```
```{r basic1-hint-2}
a <- 3
b <- 2
c <- 9

d <- (b^c)^(1/a) # initialising the variable d
d # display the value of d
```

2. Create a variable $c=a^2+b^2$ with $a=3$, $b=4$ and check if $c < 30$.

```{r basic2, exercise=TRUE, exercise.lines = 7, exercise.setup = "prepare-basic"}
a <- 3
b <- 4
# type down your codes below and click run codes to check the answer


```

```{r basic2-solution}
a <- 3
b <- 4
c <- a^2 + b^2

c # display the value of c
c > 10 #checking c < 30
```


## Vector operations

### Vector creation
We can also have arrays or vectors of values in `R`. Although there are several ways in which vector can be created, one of the most common ones is to use the `c` function. For instance, the code below creates a vector of 3 values.

```{r}
vec1 <- c(-3, 0, 2)
vec1
```

Here, `[1]` states that the answer is starting at the first element of a vector. When displaying a vector, `R` lists the elements, from left to right, using (possibly) multiple rows (depending on the width of the display).  For example:

```{r}
u <- 1:50
u
```

`R` provides a number of convenience functions for data generation. See below how one can utilise the built-in `R` functions to create vectors.


```{r}
# 5 elements of 3
rep(3, 5) 

# first 10 positive intergers
seq(1, 10)

# (1.0, 1.1, ..., 2.0)
seq(1, 2, 0.1)

# 6 equally spaced numbers between 1 and 2 (inclusive)
seq(1, 2, length = 6)
```

### Indexing vectors

We can index vectors in several ways. For instance the code below illustrates some basic indexing operations.

```{r}
vec1[1]
vec1[-1]
```
The first line extracts the first element of the vector while the second line all **but** the first element of the vector.

There are some more complicated ways to index vectors in `R`. To illustrate this code below creates a vector $vec2 = (1, 3, 4, 10, 15, 20, 50, 1, 6)$ and performs some operations:

```{r}
# initialise the vector vec2
vec2 <- c(1, 3, 4, 10, 15, 20, 50, 1, 6)
vec2

# extract the fifth, sixth and seventh elements
vec2[5:7]

# extract the first, and seventh elements
vec2[c(1,7)]

# return a logical vector that indicates whether vec2>10
vec2 > 10

# determine the positions of elements of vec2 that are larger than 10
which(vec2 > 10)

# extract the elements of vec2 that are larger than 10
vec2[vec2>10]
```

### Applying functions to a vector

We can apply different functions to a vector. Using the vector `vec2` that has been created before, perform the following operations


```{r vector3}
sqrt(vec2) # square root

log(vec2) # natural logarithm

length(vec2) # numeber of elements in a vector

sum(vec2) # sum

diff(vec2) # difference between adjacent elements

mean(vec2) # average

```


### Exercises

Use the vector `vec2` that has been created before to perform the following operations.
```{r}
vec2
```

```{r prepare-vecex}
vec2 <- c(1, 3, 4, 10, 15, 20, 50, 1, 6)
```

1. Extract the second and forth elements of the vector 

```{r vecex1,exercise=TRUE,exercise.lines=4, exercise.setup = "prepare-vecex"}


```

```{r vecex1-hint-1}
# the function c can be used to define a vector with element (2,4)

## The solutions is included in the next hint page
```

```{r vecex1-hint-2}
vec2[c(2, 4)]
```


2. Compute the sum of the average of the first five element of  the vector 

```{r vecex2,exercise=TRUE,exercise.lines=4, exercise.setup = "prepare-vecex"}


```

```{r vecex2-hint-1}
# the code 1:5 generates a sequence with the numbers from 1 to 5. 
```

```{r vecex2-hint-2}
# the function sum returns the sume of the elements of a vector

## The solutions is included in the next hint page
```

```{r vecex2-hint-3}
sum(vec2[1:5])
```

3. Extract the elements of `vec2` that are smaller or equal to 10

```{r vecex3,exercise=TRUE,exercise.lines=4, exercise.setup = "prepare-vecex"}


```

```{r vecex3-hint-1}
# remember that vec2<=10 gives a logical vector that shows whether the elements <=10 or not

## The solutions is included in the next hint page
```

```{r vecex3-hint-2}
vec3[vec3<=10]
```


## Functions

We have used so far several `R` functions. If you are unsure about what a particular function does you can get some help using either `help` or `?`. For instance,

```{r, eval=FALSE}
?sd
help(sd)
```

will display a help page for the `sd` function, with details on the input values, the output, and some examples. 

<!-- Try that by running the code below -->

<!-- ```{r help, exercise=TRUE} -->
<!-- ?sd -->
<!-- ``` -->



However, often there is no function to compute what we need, so we have to write our own functions. For instance, a function to compute the $p$-th power of $x$ can be defined as follows:

```{r}
pow <- function(x, p = 2){ 
  x ^ p
}
```


We can then use this function in several ways:
```{r}
pow(x = 2, p = 3)
pow(2, 3)
pow(1:10, 3)
```

Note that in the definition of the  function `p=2` means that `p`  is equal to 2 by default. Therefore `pow(2, 2)` returns the same value as `pow(2)`.

```{r}
pow(2, 2)
pow(2)
```

If we type `pow` in the console, the code of the function will appear:
```{r}
pow
```


### Exercise

Now complete the following code to create a function that calculates the effective annual interest rate, $i$, given the nominal annual interest rate (monthly compounding), $j$. The default value is $i=0.05$. The formula is $j=(i/12)^{12} - 1$.

```{r func,exercise=TRUE,exercise.lines=12}
j<-function(i){# note you need to specify the default value
  # type down the body of the function here
  
}

# test the output
j

temp1 <- j()
temp1

temp2 <- j(0.07)
temp2
```

```{r func-hint-1}
# to set up the default value, one should use i=0.05 instead of i
# when declaring the function

## The solutions are included in the next hint page
```

```{r func-hint-2}
# function declaration
j <- function(i = 0.05){
  j <- (1 + i/12)^12 - 1
}

# test the output
j

temp1<-j()
temp1

temp2<-j(0.07)
temp2
```


## R Packages 
A *package* is a related set of functions, including help files and data files, that have been bundled together and is shared among the `R` community. There are thousands of packages available covering different disciplines. A useful start for searching packages on a particular topic is [CRAN Task Views](https://cran.r-project.org/web/views/){target="_blank"}.

To install a package from the Internet (for instance **lifecontingencies** to perform financial and actuarial valuations) we use:

```{r, eval = FALSE}
install.packages("lifecontingencies", dependencies=TRUE)
```

For the package to be available we need to  *load* the package in `R` with the command:

```{r, message= FALSE}
library("lifecontingencies")
```

We can now see the available functions in the package:

```{r, eval = FALSE}
?lifecontingencies
```

<!-- Try that by running the code below -->

<!-- ```{r helppackage, exercise=TRUE} -->
<!-- ?lifecontingencies -->
<!-- ``` -->
