---
title: "Life Tables"
output: learnr::tutorial
bibliography: library.bib
markdown_extensions:
    - admonition  
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library("lifecontingencies")
tutorial_options(exercise.timelimit = 60)
knitr::opts_knit$set(root.dir = getwd()) 
```

## Introduction

In this tutorial we aim to discuss the implementation of life table  computations using the package `lifecontingencies` [@Spedicato2013].

The `lifecontingencies` package is the main `R` package available for performing standard financial and actuarial mathematics calculations. In this tutorial we focus only on the use of this package for performing life table calculations. If you are interested in the wider use of this package please refer to Chapter 7 of the book by @Charpentier2014.

Note that the `lifecontingencies` package has been loaded in the current `R` workspace. Therefore you can utilise the capacities of the package without explicitly loading it. However, if you are using the package outside this tutorial remember to first install the package with the code:

```{r eval=FALSE}
install.packages("lifecontingencies", dependencies = TRUE)
```

And then load it with the instruction

```{r eval=FALSE}
library("lifecontingencies")
```


## Creating a life table - Method I

In `lifecontingencies` there are several approaches for creating a life table. The first one is by inputting directly the $l_x$. 

For example, if we consider $l_0 = 120, l_1 = 119, l_2 = 118, \ldots, l_{120} = 0$, we can create a life table with the following steps:


```{r lt}
## create a life table
x <- 0:120 # create a vector X of ages, that is, integers from 0 to 120
lx <- 120:0 # create lx, that is, a vector from 120 to 0
names(lx) <- x #name each element of lx by using x

# create a life table LT1 by using the new function,
# which generates a lifetableclass object in R
LT1 <- new("lifetable", x = x, lx = lx, name="Sample life table 1")
```

We can use the `plot` function to create a plot of $l_x$ with the following code:

```{r}
# plot LT1 by using the plot() function
plot(LT1, type="l")
```

Or alternatively:
```{r}
# plot LT1 by using the plot() function with specifications
plot(LT1@x,LT1@lx, lwd = 2, col = "red", type = "l", xlab = "Age", ylab = "lx", main = "lx vs x")
```

In the previous code, note the use of `@` to access the ages in the life table (`LT1@x`) and the corresponding $l_x$ (`LT1@lx`).

```{r disdist, echo=FALSE}
question("What distribution we are assuming for the future life time T in the previous plots of lx?",
  answer("Weibull distribution",message = "The Weibull distribution is a candidate for life time models; however, in this case lx decreases linearly throught time which means T doesn't follow a Weibull distribution."),
  answer("Exponential distribution (constant force of mortality)", message = "The exponential distribution has a constant force of mortality. Since lx decreases linearly through ages, the mortality rate increases."),
  answer("The future lifetime is uniformly distributed.", correct = TRUE),
  allow_retry = TRUE
)
```


### Survival and death probabilites

Having created a life table we can easily compute survival probabilities and death probabilities with  functions `pxt`, `qxt`. 

For example, one can find the value for the survival probability, ${}_2p_{50}$, by using the following codes:

```{r}
# by using lx directly
lx["52"]/lx["50"]
# by using the pxt function
pxt(LT1,x = 50, t = 2) 
```

Similarly, one can use find out the value for the death probability, $_{10}q_{75}$, by using the following codes:

```{r}
# by using lx
1 - lx["85"]/lx["75"]
# by using the qxt function
qxt(LT1,x = 75, t = 10)
```

### Life expectancies

We can also compute life expectancies with the function `exn`. 

For example, the complete life expectancy at birth, $\mathring{e}_0$,  and the curtate life expectancy at birth, $e_0$ implied by the `LT1` life table are:

```{r}
# complete life expectancy at birth
exn(LT1, x = 0, type = "complete")

# curtate life expectancy at birth
exn(LT1, x = 0, type = "curtate")
```

```{r dislife, echo=FALSE}
question("Which one of the following statements about lifetimes and life expectancies is false?",
  answer("The curtate life expectancy computes the expected value of the interger part of the lifetime.",
         message = "Recall that curtate life expectancy is given by $e_x=E[K_x]$ with $K_x=\\lfloor{T_x}\\rfloor$."),
  answer("The complete life expectancy refers to the expected value of the lifetime.", 
         message = "Recall that complete life expectancy is given by $\\mathring{e}_x=E[T_x]$."),
  answer("The curtate lifetime is always larger or equal to the exact lifetime.", message = "The curtate lifetime is always smaller or equal to the exact lifetime. This is because the integer part of the lifetime is always smaller or equal to the exact lifetime." , correct = TRUE),
  answer("The curtate life expectancy is always smaller or equal to the complete life expectancy.",
         message = "Recall that $\\lfloor{x}\\rfloor\\leq x$ therefore $e_x=E[K_x]=E[\\lfloor{T_x}\\rfloor]\\leq E[T_x] = \\mathring{e}_x$ "),
  allow_retry = TRUE
)
```


### Exercises

Based on the life table `LT1` compute the value of: 

```{r prepare-ltprob}
x <- 0:120 
lx <- 120:0
names(lx) <- x
LT1 <- new("lifetable", x = x, lx = lx, name="Sample life table 1")
```


1. $p_{50}$ using the $l_x$ values directly.

```{r ltprob1,exercise=TRUE, exercise.lines = 4, exercise.setup = "prepare-ltprob"}

```

```{r ltprob1-solution}
lx["51"]/lx["50"]
```



2. $p_{50}$ using a built-in function of the `lifecontingencies` package

```{r ltprob2, exercise=TRUE, exercise.lines = 4, exercise.setup = "prepare-ltprob"}

```

```{r ltprob2-hint-1}
# Note that this is a one-year survival probability,
# hence one can use the pxt function.

#The solution is shown in the next hint
```

```{r ltprob2-hint-2}
pxt(LT1,x = 50, t = 1)
```



## Creating a life table - Method II

An alternative way for creating a life table is from one year death/survival probabilities using the convenience function `probs2lifetable`. To illustrate this, consider the data on one-year death probabilities from the French population in the file "FrenchTable.csv" available in the course website. We can create a life table as follows: 

```{r}
# Read the csv file
FTqx <- read.csv("FrenchTable.csv")

# Read the first few lines of the file
head(FTqx)

# Display the ages
FTqx$Age

# Create the life table
FLT <- probs2lifetable(probs = FTqx$qx, type = "qx", radix = 100000, name = "French LT")

# Create a summary of the life table
summary(FLT)
```

The `read.csv` function permits the reading of Excel data in csv format. Note that the  variable `FTqx` is a dataframe which is a data type similar in spirit to a database (named columns, distinct type). Columns in a dataframe can be accessed with the `$` symbol (e.g. `FTqx$Age`).

We can plot several quantities from the French life table:

1. A plot of the $l_x$: 

```{r}
# plot the life table 
plot(FLT, type="l")

```

2. A plot of $q_x$ both in its original scale and on the log scale: 

```{r}
# Plot of q_x vs x 
plot(FTqx$Age, FTqx$qx, type = "l", xlab="Age",ylab="qx", main = "qx vs x")
# Plot of log(q_x) vs x 
plot(FTqx$Age, log(FTqx$qx), type = "l", xlab="Age",ylab="log(qx)", main = "log(qx) vs x")
```

3. A plot of $d_x$.

```{r}
# Plot of d_x vs x
dxt(FLT, 0)
dx <- function(x) dxt(FLT, x)
dx_vec <- Vectorize(dx)
plot(FLT@x,dx_vec(FLT@x),type="l",xlab="Age",ylab="dx", main = "dx vs x")
```

In the above code we note the use of the function `Vectorize` which enables the vectorisation of scalar functions.

### Exercises

Use the French life table `FLT` to answer the following questions.

```{r prepare-french_ex}
FTqx <- read.csv("FrenchTable.csv")
FLT <- probs2lifetable(probs = FTqx$qx, type = "qx", radix = 100000, name = "French LT")
```


1. Compute ${}_{5|5}q_{60}$

```{r french_ex1,exercise=TRUE, exercise.lines = 6, exercise.setup = "prepare-french_ex"}


```

```{r french_ex1-hint-1}
# you need to compute the multiplication of two probabilities
```

```{r french_ex1-hint-2, exercise.lines = 6}
# The probabilities you need is the probability of surviving from age 60 
#to age 65, and then dying from between ages 65 and 75.

## The solution is included in the next hint page
```

```{r french_ex1-hint-3}
pxt(FLT, x = 60, t = 5) * qxt(FLT, x = 65, t = 5) 
```


2. Compute $\stackrel{\circ}{e}_{0}$ and  $\stackrel{\circ}{e}_{65}$ + 65. Which quantity is bigger and why? 


```{r french_ex2, exercise=TRUE, exercise.lines = 9, exercise.setup = "prepare-french_ex"}

```

```{r french_ex2-hint-1}
# you can use the exn function to compute the life expectancies

## The solutions are included in the next hint page
```

```{r french_ex2-hint-2}
exn(FLT, 65, type = "complete")
exn(FLT, 65, type = "complete") + 65
exn(FLT, 0, type = "complete")
```


## Assumptions for non-integer ages

The `lifecontingencies` package allows the evaluation of survival probabilities at fractional intervals, using different kinds of interpolation approaches. 

For example, using the French Life Table `FLT` we can calculate ${}_{0.6}p_{80}$ under:

- the assumption of uniform distribution of deaths (linear interpolation).

```{r}
pxt(FLT, x = 80, t = 0.6, fractional = "linear") 
```

- the assumption of constant force of mortality.

```{r}
pxt(FLT, x = 80, t = 0.6, fractional = "constant force") 

```

- the Balducci assumption (hyperbolic interpolation).

```{r}
pxt(FLT, x = 80, t = 0.6, fractional = "hyperbolic") 
```

### Exercise  

Using the French Life Table, `FLT`, calculate ${}_{0.4}q_{80}$ under the Balducci assumption.


```{r frac,exercise=TRUE, exercise.lines = 5, exercise.setup = "prepare-french_ex"}


```


```{r frac-hint-1}
# note that you should use the qxt function
# while specify the argument fractional as "hyperbolic"

## The solutions are included in the next page.
```

```{r frac-hint-2}
qxt(FLT, x = 80, t = 0.4, fractional = "hyperbolic") 
```



## References
