---
title: "Survival Analysis"
output: learnr::tutorial
bibliography: library.bib
markdown_extensions:
    - admonition  
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library("learnr")
library("lifecontingencies")
library("survival")
library("KMsurv")
tutorial_options(exercise.timelimit = 60)
knitr::opts_knit$set(root.dir = getwd()) 
```

## Introduction

This tutorial discusses the functions that are used for survival analysis in `R`, including the Kaplan-Meier and Nelson-Aelen estimators, Cox regression models and accelerated failure time models. This tutorial is based on the *Survival Analysis in R* guide available at [OpenIntro](https://www.openintro.org/stat/surv.php){target="_blank"} [@Diez2013]. 

It is assumed that the reader has a basic understanding of working in `R`. For this module, we will need the **survival** [@Therneau2015] and **KMsurv** [@Yan2012] packages. Note that both of these packages have been loaded in the current R workspace. Therefore you can utilise the capacities of the package without explicitly loading it.

However, if you are using the packages outside this tutorial remember to first install the packages with the code:

```{r eval=FALSE}
install.packages("survival", dependencies = TRUE)
install.packages("KMsurv", dependencies = TRUE)

```

And then load them with the instruction

```{r eval=FALSE}
library("survival")
library("KMsurv")
```


## KM Estimation

For the following section, we will need to use the `tongue` data set. This dataset, which is discussed in detail in @Klein2003 [Section 1.11], contains times to death for patients with cancer of the tongue according to two types of tumours. Here the `type` variable refers to the Tumor DNA profile (where 1 and 2 correspond to Aneuploid Tumor and Diploid Tumor), the `time` variable records the time to death or on-study time (in weeks), and delta is an indicator for the status of life (where 0 refers to aline and 1 refers to dead).

One can load the data by using the `data` function.

```{r}
data(tongue)
head(tongue)
attach(tongue)
```
Note the use of the `attach` function which makes the variables in `tongue` accessible from the `R` workspace.


One can create a survival object using the `Surv` function. This object will be used in the survival functions later. 

```{r}
# create a subset for just the first group by using [type==1]
cens.tongue <- Surv(time[type == 1], delta[type == 1])
cens.tongue
```

The `tongue` data correspond to **right-censored data**.  Note that the `cens.tongue` object will denote censored observations by `+`. The **survival package** also supports left-truncated right-censored data.

We want to estimate the Kaplan-Meier estimate. This is done using the `survfit` function. The `conf.int` argument is the confidence interval level while the `conf.type` is the transformation used to construct the confidence interval. By default, this latter argument is set to `log`, which is $g(t) = \log(t)$.

```{r}
fit1 <- survfit(cens.tongue ~ 1, conf.int = 0.95, conf.type = "log")
summary(fit1)
```

The Kaplan-Meier estimate can be plotted using `plot(fit1)`:

```{r KMplot}
plot(fit1, main = "KM estimate with 95% confidence intervals", xlab = "time", 
     ylab = "S(x)")
```

### Exercises

1. Calculate the KM estimate for the second group (that is, type 2) of the `alloauto` data set with 99% confidence interval. Note the `alloauto` data includes the following three variables

- time: Time to death or relapse, months
- type: Type of transplant (1=allogeneic, 2=autologous)
- delta: Leukemia-free survival indicator (0=alive without relapse, 1=dead or relapse)


```{r km, exercise=TRUE,exercise.lines=16}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the surival object for the second group
cens.alloauto <- Surv(time[type == 2], delta[type == 2])

# fit the survival curve


# print out the results


# plot the curve with confidence intervals


```

```{r km-solution}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the surival object for the second group
cens.alloauto <- Surv(time[type == 2], delta[type == 2])

# fit the survival curve
fit2 <- survfit(cens.alloauto ~ 1, conf.int = 0.99, conf.type = "log")

# print out the results
summary(fit2)

# plot the curve with confidence intervals
plot(fit2, main = "KM estimate with 99% confidence intervals", xlab = "time", 
     ylab = "S(x)")
```


2.

```{r km2, echo=FALSE}
question("Based on the results of the previous exercise, what are the KM estimate and its standard error of the survival rate at 10 months?",
  answer("0.681 and 0.0659",correct = TRUE),
  answer("0.660 and 0.0672", message = "You should look at the estimate at 8.711 instead of 10.329"),
  answer("0.531 and 0.874", message = "0.531 and 0.874 are the boundaries of the corresponding confidence interval."),
  answer("There is no event at 10 months hence the results are not available", message = "The KM estimate at 10 is the same as that of 8.711 (that is, the closet event whose time is no more than 10."),
  allow_retry = TRUE
)
```


## NA Estimation

We can also use the `survfit` function to calculate the Nelson-Aalen estimator. This is done by setting the `type` argument to `"fh"` (here `"fh"` stands for Fleming-Harrington which is an alternative name used in some textbooks for the Nelson-Aaelen estimators) as shown below.

```{r NAest}
fit2 <- survfit(cens.tongue ~ 1, conf.int = 0.95, type = "fh")
summary(fit2)
plot(fit2, main = "NA estimate with 99% confidence intervals", xlab = "time", 
     ylab = "S(x)")
```

The comparison of the Kaplan Meier and Nelson-Aalen  estimators is shown below. 

```{r KMvNA}
plot(fit1, main = "KM v.s. NA estimates", xlab = "time", ylab = "S(x)", col = "blue")
lines(fit2, col = "red")
legend("topright", c("KM", "NA"), col = c("blue", "red"), lty = 1)
```

### Exercise 

```{r kmna, echo=FALSE}
question("Does the ordering between the KM and NA lines above make sense?",
  answer("Yes. The NA estimates should be higher that KM estimates between t1 and tmax.",correct = TRUE),
  answer("No. The NA and KM estimates should be the same.", message = "Recall that we have $\\hat{S}_{KM}(t)<\\hat{S}_{NA}(t)$, for $t_1\\leq t \\leq t_{max}$."),
  answer("No. The KM estimates should be higher that the NA estimates.", message = "Recall that we have $\\hat{S}_{KM}(t)<\\hat{S}_{NA}(t)$, for $t_1\\leq t \\leq t_{max}$."),
  allow_retry = TRUE
)
```

## Log-Rank and Peto-Peto Tests

In order to conduct Log-rank and Peto-Peto tests for the difference between two survival curves, you will need to use the `survdiff` function. Setting the argument `rho` equal to zero will specify a Log-rank test. 

For instance. we can apply the Log-rank test to test if the survival times for the two types of tumours in the `tongue` dataset are the same:

```{r}
G2 <- Surv(time, delta)
logrank <- survdiff(G2 ~ type, rho = 0)
logrank 
```

Alternatively, setting the `rho` argument to 1 will specify the Peto-Peto test:

```{r}
peto <- survdiff(G2 ~ type, rho = 1)
peto 
detach(tongue)
```
The output of `survdiff` is generally self-explanatory. A $\chi^2$ statistic is computed along with a p-value. In this particular case, we cannot reject the hypothesis that both type of tumours have the same survival time.

### Exercises

1. Apply the log-rank test and test whether the survival curves for the two types of transplant of the `alloauto` data are the same. 

```{r lr, exercise=TRUE,exercise.lines=12}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the survival object 


# perform the log-rank test


```

```{r lr-solution}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the survival object 
alloauto <- Surv(time, delta)

# perform the log-rank test
lr_alloauto<- survdiff(alloauto ~ type, rho = 0)
lr_alloauto 
```

2. Apply the Peto-Peto test and test whether the survival curves for the two types of transplant of the `alloauto` data are the same. 

```{r wn, exercise=TRUE,exercise.lines=12}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the survival object 


# perform the Peto-Peto test


```

```{r wn-solution}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the survival object 
alloauto <- Surv(time, delta)

# perform the Peto-Peto test
peto_alloauto <- survdiff(alloauto ~ type, rho = 1)
peto_alloauto 
```

3. 

```{r lrwn, echo=FALSE}
question("At a critical value of 5%, what can you conclude from the test results?",
         answer("The Peto-Peto test suggests significant difference, while the log-rank test does not.", message = "Both p-values are larger than 5% so there is no significant difference."),
         answer("The log-rank test suggests significant difference, while the Peto-Peto test does not.",message = "Both p-values are larger than 5% so there is no significant difference."),
         answer("Both test do not suggest significant difference.", correct = TRUE),
         answer("Both test suggest significant difference.", message = "Both p-values are larger than 5% so there is no significant difference."),
         allow_retry = TRUE
)
```

## Cox Proportional Hazard Model

For the following examples, we will be using the `larynx` dataset. This dataset, which is discussed in detail in @Klein2003 [Section 1.8], contains times to death for male patients with laryngeal cancer. Note the data frame contains the following columns (see the R help file):

- stage: Stage of disease (1=stage 1, 2=stage2, 3=stage 3, 4=stage 4)

- time: Time to death or on-study time, months

- age: Age at diagnosis of larynx cancer

- diagyr: Year of diagnosis of larynx cancer

- delta: Death indicator (0=alive, 1=dead)

We first conduct a similar procedure as in the case of fitting the KM/NA estimators and create a survival object from the data.

```{r}
data(larynx)
head(larynx)
attach(larynx)
cens.larynx <- Surv(time,delta)
cens.larynx
```

The Cox PH model can be fitted  using the `coxph` function, with the argument `method` specifying the estimation procedure to be used. Note the use of the `as.factor` function to allow each of the different values of the variable "stage" to be treated separately in the regression. We can also obtain the baseline hazard function using the `basehaz` function. This is demonstrated below.

```{r}
Cox.Reg <- coxph(cens.larynx ~ as.factor(stage) + age, method = "breslow")
summary(Cox.Reg)
```


We can also obtain the baseline hazard function using the `basehaz` function. This is demonstrated below.

```{r}
Base.H <- basehaz(Cox.Reg, centered = FALSE)
plot(Base.H$time, Base.H$hazard, xlab = "t", ylab = "H_0(t)", 
     main = "Baseline Hazard Rate", type = "s")
```

### Exercises
Based on the output of the previous Cox regression, answer the following questions. 

```{r coxreg2, echo=FALSE}
question("Which one of the following statements is correct?",
         answer("All the covaraites turn out to be significant (based on a 5% critical level)", message = "The p-values for stage 2, stage 3 and age are all larger than 0.05."),
         answer("The likelihood ratio and Wald tests show that it is worth including the covariates compared to the baseline model.",correct = TRUE),
         answer("The basedline model correspond to stage 1 and age 1",message = "The baseline age is 0."),
         allow_retry = TRUE
)
```

```{r coxreg3, echo=FALSE}
question("Suppose that A is one year elder than B and both of them are in stage 2. What is the ratio of mortality rate between A and B (approximately) given the numerical output of the fitting? Does this change if they are both at stage 3 (assuming that the Cox model is correct)?",
         answer("0.01908, Yes", message = "The ratio should be 1.01908 (which is exp(coef) or 1+coef approximately). Furthermore, the ratio stays the same due to the Cox regression assumption of proportionality."),
         answer("1.01908, Yes", message = "The ratio stays the same due to the Cox regression assumption of proportionality."),
         answer("0.01908, No", message = "The ratio should be 1.01908 (which is exp(coef) or 1+coef approximately)."),
         answer("1.01908, No", correct = TRUE),
         allow_retry = TRUE
)
```

```{r coxreg4, echo=FALSE}
question("Which of the following transition will lead to the highest percentage change of mortality rate (given the same age)?",
         answer("Transition from stage 1 to stage 2.", message = "The increase of the mortality rate is 14.862%, which is not the highest."),
         answer("Transition from stage 2 to stage 3.", message = "The increase of the mortality rate is 64.837%, which is not the highest."),
         answer("Transition from stage 3 to stage 4.", correct = TRUE),
         allow_retry = TRUE
)
```

## Accelerate Failure Time model

Finally, we can fit in a similar manner the appropriate AFT model by specifying the distribution in the `dist` argument of the `survreg` function. We demonstrate the use of this function below for a Weibull AFT model.   

```{r}
Weib.Reg <- survreg(cens.larynx ~ as.factor(stage) + age, dist = 'weibull')
summary(Weib.Reg)
```
In the output, `(Intercept)` and `Log(scale)` correspond to estimates of $\mu$ and $\log \sigma$.

### Exercise

Apply the Weibull AFT model to the `alloauto` dataset with the covariate type (as a factor)


```{r aft, exercise=TRUE,exercise.lines=12}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the surival object 
alloauto <- Surv(time,delta)

# apply the Weibull AFT model


```

```{r aft-solution}
# load and attach the data
data(alloauto)
attach(alloauto)

# create the surival object 
alloauto <- Surv(time,delta)

# apply the Weibull AFT model
Weib.alloauto <- survreg(alloauto ~ as.factor(type), dist = 'weibull')
summary(Weib.alloauto)
```

```{r aft2, echo=FALSE}
question("Which one of the following coefficient is significant (at a critical value of 5%)?",
         answer("Stage 2", message = "The p-value is higher than 5% hence it is not significant."),
         answer("Stage 3", message = "The p-value is higher than 5% hence it is not significant."),  
         answer("Stage 4", correct = TRUE),         
         answer("age", message = "The p-value is higher than 5% hence it is not significant."),
         allow_retry = TRUE
)
```


```{r aft3, echo=FALSE}
question("Suppose that A is one year older than B and both of them are in stage 2. What is the ratio of mortality rate between A and B (approximately) given the numerical output of the AFT fitting?",
         answer("1.01908", message = "This is calculated based on the fitting of the Cox regressoin model."),
         answer("0.9826522", message = "Note that you should change the sign of the estimated coefficient before taking the exponential."),
         answer("0.0175", message = "The ratio should be exp(0.0175) or approximately 1+the coefficient."),
         answer("1.017654", correct = TRUE),
         allow_retry = TRUE
)
```


## References

